---
title: "Data 607 Spring 2019 HW 5"
author: "Eunice Ok"
date: "3/3/2019"
output: html_document
---

```{r setup, include=FALSE}
require(tidyr)
require(dplyr)
library(ggplot2)
library(ggthemes)
```



#### READ IN DATA 
```{r}
file = "/Users/euniceok/PycharmProjects/cuny/spring2019/Week5/data/airline_delays.csv"
df = read.csv(file, na.strings=c(""," "))
df
```

#### TIDY and TRANSFORM DATA 
```{r}
df <- df %>% 
  # convert wide dataset to long on all cols except X and X.1
  gather(key, value, -X, -X.1, na.rm=TRUE) %>% 
  # rename all cols to be intuitive
  rename(city = key, counts = value, airline = X, status = X.1) %>% 
  # fill out missing airline names
  fill(airline) 
head(df)

# another version of tidy df for easier analysis
tmp <- df %>% 
  spread(status, counts) %>% 
  rename(ontime = 'on time') %>% 
  mutate(total = delayed + ontime)
tmp

```

#### ANALYZE DATA

1. calculate delay and ontime rates for each airline
```{r}
# what is the relative rate of delayed flights for each airline? 
airline_delay <- tmp %>%
  group_by(airline) %>%
  summarise(delays = sum(delayed),totals = sum(total)) %>%
  mutate(del_rate = delays / totals)

# what is the on time rate for flights for each airline? 
airline_ontime <- tmp %>%
  group_by(airline) %>%
  summarise(ontime = sum(ontime),totals = sum(total)) %>%
  mutate(ontime_rate = ontime / totals)

airline_tbl <- inner_join(airline_delay, airline_ontime, by="airline")
airline_tbl %>% select(airline,del_rate,ontime_rate)
airline_tbl
```

```{r}
atplot <- airline_tbl %>% 
    select(airline,delays, ontime) %>%
    gather(delays, ontime, -airline) %>%
    rename(status = delays, flightcts = ontime)  
atplot
```


2. plot delay and ontime flight counts for each airline
```{r}
g <- ggplot(atplot, aes(airline, flightcts))
g + geom_bar(aes(fill=status), width = 0.5, stat="identity") + 
  theme(axis.text.x = element_text(angle=65, vjust=0.6)) + theme_tufte()
```


3. calculate delay and ontime rates for each airline and city
```{r, include=FALSE}
#  delay rate for each airline-city? 
aircitydelay <- tmp %>%
  group_by(airline, city) %>%
  summarise(delays = sum(delayed),totals = sum(total)) %>%
  mutate(del_rate = delays / totals)

#  ontime rate for each airline-city? 
aircityontime <- tmp %>%
  group_by(airline, city) %>%
  summarise(ontime = sum(ontime),totals = sum(total)) %>%
  mutate(ontime_rate = ontime / totals)

aircttbl <- inner_join(aircitydelay, aircityontime, by= c("airline" = "airline","city"="city"))
aircttbl %>% select(airline, city, delays, ontime, del_rate, ontime_rate)

```


4. plot rates for each airline and city

```{r}

# ALASKA
alctplot <- 
  aircttbl %>% 
  filter(airline == "ALASKA") %>% 
  ungroup() %>%
  select(city, del_rate, ontime_rate) %>%
  gather(del_rate, ontime_rate, -city) %>%
  rename(status = del_rate, flightcts = ontime_rate) 

# AM WEST
amctplot <- 
  aircttbl %>% 
  filter(airline == "AM WEST") %>% 
  ungroup() %>%
  select(city, del_rate, ontime_rate) %>%
  gather(del_rate, ontime_rate, -city) %>%
  rename(status = del_rate, flightcts = ontime_rate) 
```

ALASKA PLOT
```{r}
ggplot(alctplot,aes(x=city,y=flightcts,fill=factor(status)))+
  geom_bar(stat="identity",position="dodge")+
#  scale_fill_discrete(name="status",
#                      breaks=c(1, 2),
#                      labels=c("delay", "ontime"))+
  xlab("airline")+ylab("rates") + theme_tufte()
```

AM WEST PLOT
```{r}
ggplot(amctplot,aes(x=city,y=flightcts,fill=factor(status)))+
  geom_bar(stat="identity",position="dodge")+
#  scale_fill_discrete(name="status",
#                      breaks=c(1, 2),
#                      labels=c("delay", "ontime"))+
  xlab("airline")+ylab("rates") + theme_tufte()
```


```{r}
delctplot <- aircttbl %>%
  select(airline, city, del_rate)
delctplot
```
```{r}
ggplot(delctplot,aes(x=city,y=del_rate,fill=factor(airline)))+
  geom_bar(stat="identity",position="dodge")+
  #scale_fill_discrete(name="status",
  #                    breaks=c(1, 2),
  #                    labels=c("alaska", "am west"))+
  xlab("city")+ylab("rates") + theme_tufte()
```

#### FINDINGS
- Because these airlines may have different overall volume of flights, I converted the data into delay/on time rates.

- Delay rates were more useful, and since they were the inverse of on time rates, I used delay rates as the main metric for comparison. 

- Overall, the delay trends for each city were similar between both airlines. 

- Both airlines had the highest delay rates in San Francisco and lowest delay rates in Phoenix. 

- This indicates that the primary cause of the delay is related to the location as opposed to the airline. 

- However, the magnitude of the delay in each differed between the two airlines.

- Interestingly, at a high level across all cities, Alaska had a higher delay rate. But within each city, AM West had higher delay rates. Perhaps this is an example of Simpson's Paradox. 

- The biggest gap is in San Francisco, where AM West has a delay rate 12 percentage points higher than Alaska. 

- There is also a sizeable gap in Seattle (9 pct pts), which indicates that AM West may have some inefficiencies - whether it is due to slower operations, equipment, etc that is an opportunity for AM West to address to catch up to Alaska.  




